---
title: "Project Report"
author: "Kyle Conrad and Sarah Baker"
date: "11/19/2023"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Questions of interest
\begin{itemize}
  \item How is the growth of plants within the Sonoran Desert affected by different fertilization methods and nitrogen atmospheric deposition.
  \item Is there a difference in the impact of these fertilization methods and nitrogen deposition between the three zones studied? These zones are: with the urban core, west of the urban core, and east of the urban core.
  \item Is there a difference apparent between affect on growth of annual plants as compared to perennial plants? What is the impact on growth of Larrea Tridentata plants specifically?
\end{itemize}

# Analysis and Data Processing
## Preprocessing the Data
Loading the required packages:
```{r}
library(tidyverse) # For re-formatting data sets
```

Read in data:
```{r}
annuals_biomass <- read.csv("Data/632_annuals_biomass.csv")
```

Prior to beginning analysis, the data should be cleaned to remove justifiable outlying observations. Additionally, variables which will not be used will be removed from each sub data set prior to combining data.\

Begin by cleaning up `annuals_biomass` which contains the response variable of interest, biomass of annuals.
```{r, results="hide"}
# Check number of missing subquad_orientation measurements
sum(is.na(annuals_biomass$subquad_orientation))

# Drop subquad_orientation from data set due to high number of missing observations
annuals_biomass <- subset(annuals_biomass, select = -c(subquad_orientation))

# Remove uninformative notes variable values
unique(annuals_biomass$notes)

vec <- c("average bag wt (n=10) = 6.733", "changed from 17.34 to 27.34 after rechecking", "2 bags combined for one sample", "average bag wt (n=10) = 1.437", "Some of the brown bags differed in type due to location where the bags were purchased. The two types of bag used were lunch bags and giant lunch bags. The lunch bags were smaller and the giant lunch bags were larger. Average weight of small bag (n = 10) = 7.297; average weight of large bag (n = 10) = 10.241.", "very little vegetation", "bubble in compass", "Ambrosia", "B4 on bag", "Dry plants")

annuals_biomass$notes[annuals_biomass$notes %in% vec] <- NA

# Check the notes variables and set mass to 0 when no biomass was collected, per notes variable
vec <- c("no biomass to collect", "no sample; average bag wt (n=10) = 6.733", "no biomass to collect in sample quad; average bag wt (n=10) = 1.437", "no plants to collect")

annuals_biomass$mass[annuals_biomass$notes %in% vec] <- 0
annuals_biomass$notes[annuals_biomass$notes %in% vec] <- NA

# Remaining observations have unresolved sampling errors therefore these observations are dropped.
vec <- unique(annuals_biomass$notes)
vec <- vec[!is.na(vec)]
annuals_biomass <- subset(annuals_biomass, !(notes %in% vec))

# Drop notes variable
annuals_biomass <- subset(annuals_biomass, select=-c(notes))

# Samples are collected in the spring each year and sampled once a year, therefore, date variable contains no useful information for this analysis. Drop the date variable
annuals_biomass <- subset(annuals_biomass, select=-c(date))

# If mass is NA for any other observations then will remove those observations. Cannot determine reasoning behind the NA's.
annuals_biomass <- subset(annuals_biomass, !is.na(mass))

# plot_id contains the same information as treatment_code, therefore it will be dropped
annuals_biomass <- subset(annuals_biomass, select = -c(plot_id))

# Drop all obserations where mass is negative, there is no useful info to be gleamed by them
annuals_biomass <- annuals_biomass[annuals_biomass$mass >= 0,]
```

Finally, a variable must be added to group the different sites by their common geographic region. That is whether the region is west, within, or east of the city core. 
```{r}
annuals_biomass$region <- NA
annuals_biomass$region[annuals_biomass$site_code %in% c("DBG", "MVP", "PWP", "SME", "SMW")] <- "core"
annuals_biomass$region[annuals_biomass$site_code %in% c("EME", "EMW", "SNE", "SNW", "WTM")] <- "west"
annuals_biomass$region[annuals_biomass$site_code %in% c("LDP", "MCN", "MCS", "SRR", "UMP")] <- "east"
```

## Analysis
Load packages to be used for analysis and set seed for replication purposes:
```{r, results='hide', warning=FALSE}
library("rstan")
library("bayesplot")
library("ggplot2")
```

```{r, echo=FALSE}
# Setting options of rstan such that a compiled copy of the code is saved to the disk, save on compilation time
rstan_options(auto_write = TRUE)
```

Apply the log transformation to the response variable and formatting the data for the rest of analysis. Note some observed masses were zero, therefore a small constant value was added to each mass observation prior to the log transformation. Additionally, year will be coded as the number of years since 2008.
```{r}
X <- subset(annuals_biomass, select=-c(mass))
X$year <- X$year - 2008
Y <- log(annuals_biomass$mass + .0001)
```

Now, inspect the distribution of the response to investigate the shape of the distribution. 
```{r}
plot(density(Y))
```

Notice that there is a peak in the density of the responses at Ln($10^{-4}$) which does not follow the trend of the other observations in the data set. This corresponds to the zero values in our original data and suggests that when extremely sparse growth was observed by those collecting biomass that the decision was made to simply log that as no growth. These outliers will skew the results of our analysis and thus the decision was made to remove these observations.
```{r}
annuals_biomass <- annuals_biomass[annuals_biomass$mass > 0,]

X <- subset(annuals_biomass, select=-c(mass))
X$year <- X$year - 2008
Y <- log(annuals_biomass$mass)
```

Note, it is no longer need to add a small constant to enable log transformation of the response.\

Again, inspect the distribution of the transformed response.
```{r}
plot(density(Y))
```

\newpage

LEFT BLANK INTENTIONALLY

\newpage

Create the design matrix for the chosen model for import into stan. Note, subplot is dropped since it is assumed to have a negligible effect on the model's accuracy.
```{r}
X <- data.frame(model.matrix(mass ~ . - subplot - site_code - region, annuals_biomass))
colnames(X)[1] <- "Offset"

#Standardize to speed up stan sampling, note binary variables are standardized as suggested in section 4.2 of Gelman and Hill (http://www.stat.columbia.edu/~gelman/arm/)
X$year <- (X$year - mean(X$year)) / sd(X$year)
X$treatment_codeN <- (X$treatment_codeN - mean(X$treatment_codeN)) / (2*sd(X$treatment_codeN))
X$treatment_codeP <- (X$treatment_codeP - mean(X$treatment_codeP)) / (2*sd(X$treatment_codeP))
X$treatment_codeNP <- (X$treatment_codeNP - mean(X$treatment_codeNP)) / (2*sd(X$treatment_codeNP))
X$location_within_plotP <- (X$location_within_plotP - mean(X$location_within_plotP)) / (2*sd(X$location_within_plotP))
Y <- (Y - mean(Y))/sd(Y)

# Modify dataset such that it may be read by stan
annuals_biomass$region_int <- as.integer(as.factor(annuals_biomass$region))
annuals_biomass$site_code_int <- as.integer(as.factor(annuals_biomass$site_code))

# Set the site regions as an array corresponding to the site coding above
site_region <- rep(NA, length(unique(annuals_biomass$site_code)))
for (i in 1:length(site_region)) {
  site_region[i] <- annuals_biomass$region_int[annuals_biomass$site_code_int == i][1]
}
```

STAN code using non-centered model:
```{r}
noncent_hier_code <- "
data {
  int<lower=0> N;               // num individuals
  int<lower=1> K;               // num ind predictors
  int<lower=1> J;               // num sites
  int<lower=1> L;               // num site predictors
  int<lower=1> I;               // num regions
  int<lower=1,upper=J> jj[N];   // site for individual
  int<lower=1,upper=I> ii[J];   // region for site
  matrix[N, K] x;               // individual predictors
  row_vector[L] u[J];           // group predictors
  vector[N] y;                  // outcomes
}

parameters {
  vector[K] alpha[J];           // indiv coeffs by group (reparam)
  matrix[L, K] gamma[I];           // group coeffs
  corr_matrix[K] Omega;         // prior correlation
  vector<lower=0>[K] tau;       // prior scale
  real<lower=0> sigma;          // prediction error scale
  real mu;                      // prior mean for region
  real<lower=0> eta;            // prior std for region
}

transformed parameters {
  vector[K] beta[J];            // Individual Coeffs by group
  for (j in 1:J)
    beta[j] = to_vector(u[j] * gamma[ii[j]]) + (quad_form_diag(Omega, tau) * alpha[j]);
}

model {
  mu ~ normal(0, 2);
  eta ~ inv_gamma(.1, .1);

  for (i in 1:I) {
    to_vector(gamma[i]) ~ normal(mu, eta);
  }

  tau ~ cauchy(0, 2.5);
  Omega ~ lkj_corr(2);
  
  sigma ~ inv_gamma(.1, .1);
  
  for (j in 1:J) {
    alpha[j] ~ normal(0,1);
  }
  
  {
    vector[N] x_beta_jj;
    for (n in 1:N)
      x_beta_jj[n] = x[n] * beta[jj[n]];
    y ~ normal(x_beta_jj, sigma);
  }
}

generated quantities {
  real y_rep[N];
  {
    vector[N] x_beta_jj;
    for (n in 1:N)
      x_beta_jj[n] = x[n] * beta[jj[n]];
    y_rep = normal_rng(x_beta_jj, sigma);
  }
}
"

noncent_hier <- stan_model(model_code = noncent_hier_code)
```

```{r}
hier_data <- list(
  N = nrow(X), 
  K = ncol(X), 
  J = length(unique(annuals_biomass$site_code)), 
  L = 1, 
  I = max(site_region),
  jj = annuals_biomass$site_code_int, 
  ii = site_region,
  x = X, 
  u = matrix(rep(1, length(unique(annuals_biomass$site_code))), nrow = length(unique(annuals_biomass$site_code))), 
  y = Y
  )

nfit <- sampling(noncent_hier, data = hier_data, iter= 3000, warmup = 1500, chains=1)
```

```{r}
post_smp <- as.data.frame(nfit)
post_smp <- post_smp[,-(1:48)]
colnames(post_smp)[1:6] <- colnames(X)  

plot(density(post_smp$Offset))

```

\newpage

# Appendix
## Methods and Sampling Descriptions
### Fertilizer Application
Fertilization with N, P, and N+P of 20 m x 20 m plots at 4 sites each in upwind, core, and downwind positions will be done twice per year to test the hypothesis that urban N deposition pushes the desert system toward P limitation. The level of N addition (as NH4NO3) is chosen to be ~ twice the current maximum deposition rate, i.e., 60 kg kg ha-1 y-1; P addition level is 1/5 (by mass; 1/11 by atoms) that of N at 12 kg ha-1 y-1. Fertilizer is applied to treatment plots each winter and summer to coincide with the bimodal precipitation patterns characteristic of the region. The prescribed weight of dry, granular fertilizer (ammonium-nitrate for nitrogen treatments (N, N+P), triple-super phosphate for phosphorus treatments (P, N+P)) is applied using a hand spreader. As of August 2015, the rate of N fertilizer application was reduced to 30 kg ha-1 y-1, with a resulting N:P ratio of 2.5:1.

### Plant Root Simulator
Soil anion and cation availability are quantified twice annually using ion-exchange membranes. Ion-exchange membranes (Plant Root Simulator (PRS) probes) are provided and analyzed by Western Ag Innovations, Saskatchewan, Canada. Probes for the analysis of soil anions and cations are deployed in pairs at locations under the canopy of L. tridentata (n=4) and in inter-plant spaces (n=4) within selected study plots. The probes are inserted vertically into the top 15 cm of soil, and left undisturbed until extraction after approximately 6 weeks in the ground. Upon extraction, the 8 probes of each type (anion/cation) from each habitat (plant/inter-plant) are pooled, washed with distilled water in accordance with protocols outlined by Western Ag Innovations Inc., and shipped overnight to Western Ag Innovations Inc. for analysis. PRS probe analyses typically focus on nitrogen species (i.e., NO3-N and NH4-N). However, a broader suite of analytes is assessed approximately every 3-5 years.

### Stem Growth
Stem length on select L. tridentata shrubs are measured twice per year—once in the fall and again in the spring. Five L. tridentata shrubs (designated L1 thru L5) have been selected as study plants in each plot at all CN Dep sites. Measurements are taken at 4 points on each study plant—one on a stem facing each of the cardinal directions. Each time the shrubs are measured, a new piece of colored lab tape is placed around the stem, which will be used as the point from which to measure stem elongation during the following period. Field Protocol: Before arriving at the study site, cut ca. 1-inch pieces of lab tape in half length-wise and loosely stick to back of clipboard. Make enough pieces for each direction on each plant (20 per plot). Label each piece with a plant number (L1-L5) on the left side of the tape and a cardinal direction (N, S, E, W) on the right side. Study plants are marked L1 thru L5 with a silver tree tag. For each of the 5 study plants per plot: 1. Measure stem elongation: a. Look for a piece of colored tape in each cardinal direction of the shrub. Make sure the calipers are calibrated to zero and you are measuring in mm. Measure the stem (in mm) from the node directly above the piece of tape to the last node on the stem. b. If the stem branches at any point above the tape, measure each branch separately and record all the branch lengths separated by commas on the datasheet. c. Do not measure the stalks of flowers or fruits. d. If the leaves directly above or below a node have fallen off and the stem is not measurable, make a note on the datasheet. e. If the tape cannot be found, check on the ground around the shrub to see if the tape has fallen off or the branch has died. Note on the datasheet if the tape is missing or the branch is dead. Do not leave any blank spaces on the datasheet! f. Remove the old tape from the stem. 2.Measure re-taped stem lengths: a. Re-tape the stem with a new piece of tape. Place the tape close to the end of the stem, but with a measurable internode between the tape and the end of the stem. b. Measure the re-taped stem lengths as above and record on the datasheet.

### Leaf Tissue ICPMS
Field collection: (1) Randomly choose 5 creosote plants in each of the 5 study plots for every site (15 sites total). Do not choose shrubs chosen for stem-elongation (growth) measurements. (2) From each plant, randomly select one stem in each of 4 cardinal directions. (3) From each stem, pick 4, fully developed leaf pairs from the green section of the primary stem. We are defining fully developed leaves as those that are just behind the newest leaves (the newest leaves are the 2 undeveloped green ones plus the ones just attached to them). Do not choose leaves that are farther back down the stem or leaves that are brown and/or damaged. (4) Place all leaves from each plot in a coin envelope. (5) Place coin envelopes into a dryer (60 degrees C) upon returning to the laboratory or within 48 h. Lab processing: (1) Rinse all picked leaves with DI water in a clean fine meshed sieve and place in a coin envelope with the site name, plot name, and date processed. (2) Dry all leaves in coin envelopes at 70 degrees C for at least 48 h before grinding. (3) Fill polycarb vial approximately 1/2 full with leaf litter, add one of the polycarb balls and cap vial, place vials in ball mill chamber and mill for 10 minutes. ICPMS: Dried and milled L. tridentata leaf tissue is microwave digested using a CEM MARSXpress. Approximately 0.2500 g of sample was allowed to react with 10 mL of Trace Metal Grade nitric acid (Fisher #A509-P212) for 15 minutes, uncapped, then ran on the CEM using the following method: 1600W at 100% power, ramped up to 180 degrees C over the course of 15 minutes, and held at 180 degrees C for 15 minutes. Samples were diluted at 100X prior to being run by ICP-MS. Samples were analyzed using a Thermo Fisher Scientific X Series 2 quadrapole ICP-MS and Cetac ASX- 520 autosampler. Sample introduction consisted of a conical spray chamber with impact bead and concentric nebulizer with a flow of ~1 ml/min. The spray chamber was cooled to 3 degrees C by a Peltier cooling system. Collision Cell Technology (CCT) mode was utilized to reduce interferences by the argon gas used to generate the plasma by using a mixture of 7% hydrogen and 93% helium for the following elements: Fe, Se, As, and Ca. All other elements were run in normal mode. Internal standards (Sc, Ge, Y, In, and Bi) were used to compensate for sample matrix effects and instrument variability.

### Annuals Biomass
Each spring when there is sufficient rainfall, we estimate aboveground net primary productivity (ANPP) of herbaceous annual plants in each of the Desert Fertilization Experiment study plots by harvesting the aboveground portion of annual herbs from a 1/2 m x 1/2 m section of established 1 m x 1 m sub-plots. These plots are distinguished from similar 1 m x 1 m sub-plots that are used for community composition. Take care to avoid the community composition sub-plots! Only harvest the sub-plots that are designated for biomass. Protocol overview: In each Desert Fertilization experiment plot, we established 4 biomass sub-plots per plot, two in each of two patch types, between shrubs (IP for inter-plant space) and under the canopies of L. tridentata (P for under plant). Each sub-plot is coded with the patch type (IP or P); the letter B for biomass (to distinguish it from the community composition sub-plots, which are not harvested and do not have a letter B after the patch type code); and the sub-plot number (1 or 2). For example, IPB1 & IPB2 are interplant space sub=plots one and two, and PB1 & PB2 are sub-plots 1 and 2 under L. tridentata shrubs. Remember that the biomass sub-plots are distinguished from the community composition sub-plots with the letter B after the patch type code (example, IPB1 = Biomass sub-plot 1, inter-plant space). Within each 1 m x 1 m biomass sub-plot, teams will clip aboveground annual biomass from one 1/2 m x 1/2 m quadrat in each 1-meter biomass sub-plot. This sub-section is rotated each year so that any given patch of ground is harvested at the most once every four years. Teams will carefully write the plot identifier on paper bags. (Double-check the labels! Labeling mistakes are the most common error in this procedure). Then teams will cut all green and brown herbaceous plant biomass that occurs within the boundaries of the 1/2 m x 1/2 m sub section and place it in the pre-labeled bag. After the harvest is complete, teams will dry the bags and biomass at 60 degrees C for 48 hr then weigh the biomass (subtracting the weight of the bag) on a scale and record.

### Annuals Composition
Overview: In spring 2008 within each site, the CAP LTER randomly chose and permanently marked four 1 m x 1 m sub-plots within each 20 m x 20 m main plot, split between two patch types, under L. tridentata shrubs and inter-plant spaces (hereafter, under L. tridentata, P and Inter-plant space; IP). At peak herbaceous biomass (in approximately March of each year), percent cover of all annual herbaceous species rooted within these four plots (two each under L. tridentata and in interplant spaces i.e. between shrubs) will be estimated. Note that there are also four 1 m x 1 m Biomass sub-plots in each 20 m x 20 m plot – these are different than the community composition plots described here because they are clipped each year while the community composition plots are not. Care should be taken in the field to identify the correct four plots for community composition and avoid the four plots to be harvested for biomass. Steps: 1. On the top of the data sheet (See Table 3 for sample), record the date; names of the sampling crew (i.e. plant identifier, recorder); site (three letter code); plot number (1 to 75); treatment (C1 – no sampling from C2 plots, Nitrogen [N], Phosphorus [P], and Nitrogen+Phosphorus [NP]). Double check the plot numbers and sub-plot numbers! 2. All observers should first calibrate their percent cover estimates on several test plots in the field prior to percent cover measurements. You want to make sure all botanists are seeing the plots similarly. 3. Place the 1 m x 1 m quadrat over the corner rebar at each 1-m2 plot. Working in teams of two (one recorder, one plant identifier), proceed through the steps below. 4. Estimate and record (at the bottom of the data sheet) the relative areal cover of plant matter (e.g. green material) versus bare ground (e.g. brown material) at the soil surface in the plot (e.g. 70% green, 30% brown). If there are soil crusts, measure those as well in a separate category. 5. Identify which annual species are present in the plot and mark a small dot in the corner of the datasheet to easily find later. Use roughly the same amount of effort in each plot to find plant species. As you get better, it will take less time because you are getting more skilled; as you get tired, it will take more time; in simpler systems, it will take less time. There is no a priori way to know when you are using the same amount of effort, but this is usually approximated by time (in minutes: typically 1-3 minutes per species). Each botanist should do some test plots to see how long it takes him/her to find all the species in the plot. Different botanists will take different amounts of time to apply the same effort– this is okay – but the same botanist should use approximately the same amount of time per plot. 6. Next, estimate and record the cover of each species individually to the nearest whole number (e.g., 1%, 2%, 3%…97%, 98%, 99%, 100%). If a species is present in a plot, but does not cover a full 1%, record it as <1%. The sum of all individual species cover can be greater than 100% since some species overlap. 7. Finally, estimate and record the cover of the overstory shrubs within the plot or overhanging the plot. We estimate cover for all annuals rooted in the quadrat, but areal cover of any shrubs rooted in or hanging over the plot. For IP plots, this will be zero most of the time (i.e. unless there is a large shrub nearby that has grown over the plot). For P plots, this will be have two components: (1) the areal projection of the shrub onto the 2-D grid, and (2) the area of the stem base or any other shrub stems. It is expected that the cover of the shrubs overlap the herbaceous species. However if two shrub species overlap in their cover aggregate these into a single larger individual (rare case). 8. Unknown species are recorded with unique identifiers (e.g., Unknown 1, Unknown 2, etc.) along with a visual description of the plant and a picture to be used to help identify the unknown in the lab. When possible, a sample of the unknown species should be collected from outside of the plot in a labeled coin envelope and brought to the herbarium to be identified. Unknown species are usually very small and or undeveloped, and may not be able to be identified to species – this is okay. Err on the side of lumping unknown species into other known species in the same plot. 9. Use caution not to include the percent cover of desiccated/dead plants within the plots from the previous season (these can usually be identified easily as white, old plant material). However, depending on the timing of data collection, some herbaceous plants from the current growing season may have passed peak biomass and already begun to desiccate (dry out or senesce). In this case, estimate the percent cover of these plants as if they were at peak biomass. 10. Before you leave a site, confirm that you have datasheets for all plots and sub-plots, and that there are no duplicates or ambiguous plot labels on your sheet.

### Measuring Bulk and Throughfall NO3 and NH4 Deposition
Ion exchange resin (IER) collectors are used to measure bulk (wet) deposition in interplant open spaces and throughfall (wet and dry) deposition under the dominant shrub. Both bulk and throughfall collectors are deployed with 2 replicates at each site. Collectors are deployed continuously over four 3-month sampling intervals per year – approximately January to March, March to June, June to September, and September to December. Collectors are in the field for an average deployment of 89 days.

Bulk and throughfall IER collectors are constructed with a modified design following Fenn and Poth (2004) and Simkin and colleagues (2004). For each sampling period, IER collectors are made with hydrochloric acid-washed supplies. Each IER collector consisted of a 12-inch PVC pipe filled with 60 mL Monosphere Dowex Resin (Dow Chemical Company). Due to supply issues, Amerlite IRN-150 ion exchange resin replaced Monosphere Dowex in spring 2019. The resin-filled pipe was closed on one end with a PVC cap with 5 – 7 holes drilled for drainage and filled with glass wool to prevent resin loss. On the other end of the resin-filled PVC pipe, nitex screening is used to prevent debris in the resin and a plastic funnel (open area = 314 cm2) is placed onto the top. The funnel is covered with mesh and a bird spikes to prevent debris contamination of the resin.

In the field, bulk collectors are installed 1.5 meters above the ground on rebar in open space areas without canopy cover. Throughfall collectors are placed under L. tridentata, where 2-ft holes were dug and filled partly with aquarium rocks to allow for drainage during storms. All IER collectors (including the funnels and the L. tridentata branches for throughfall collectors) are rinsed with 500 mL di-ionized water before being collected at the end of the sampling period. In addition, field blanks are prepared without a funnel and instead capped on both ends. Field blanks are deployed at one urban (DBG) and two outlying sites (WTM and LDP) with three replicates at each site for both bulk and throughfall field blanks.

After exposed samplers are collected in the field, NH4 and NO3 ions are extracted from each resin sample with 200 mL 2M potassium chloride (KCl) solution. KCl-resin slurries are shaken for one hour and then filtered through Whatman 42 filters pre-leached with KCl. In addition, three KCl extract blanks are prepared. All KCl extracts are analyzed on the continuous flow injection Lachat QuikChem 8000 (Lachat Instruments) for NH4 and NO3.

## Descriptions of Data Sets and Variables
### Annuals_Biomass
Biomass (g) of annual plants harvested from subplots within Desert Fertilization study plots. One-meter subplots include locations around a Larrea tridentata plant and locations in the interplant space between shrubs. Material is harvested from 0.25 square meter quadrants within each subplot. All harvests occur during the spring.
\begin{itemize}
  \item site\_code - site name abbreviation
  \item plot\_id - plot id number
  \item treatment\_code - treatment name abbreviation
  \item location\_within\_plot - habitat within plot where probes were deployed
  \item subplot - subplot number (one of two replicates)
  \item subquad\_orientation - direction of sub-quadrant from which biomass was harvested
  \item date - date of data collection
  \item year - year of data collection
  \item mass - mass of biomass harvested in 0.25 meter Square quadrant
  \item notes - field and processing notes
\end{itemize}

### Annuals_Composition
Composition of annual plants and some other characteristics (e.g., bare soil, base or canopy of perennial plants) at subplots within Desert Fertilization study plots. One-meter subplots include locations around a Larrea tridentata plant and locations in the interplant space between shrubs. All measurements collected in the spring.
\begin{itemize}
  \item site\_code - site name abbreviation
  \item plot\_id - plot id number
  \item treatment\_code - treatment name abbreviation
  \item location\_within\_plot - habitat within plot where probes were deployed
  \item subplot - subplot number (one of two replicates)
  \item date - date of data collection
  \item year - year of data collection
  \item collector - initials of subplot surveyors
  \item cover\_type - taxon of annual plant, plot characteristic, or logistical note
  \item cover\_category - nature of the observation (taxonomic, general plot characteristic, logistical)
  \item cover\_amount - proportion of subplot as a decimal fraction of this type
\end{itemize}

### Atmospheric_Deposition
Ammonium-nitrogen and nitrate-nitrogen as measured by ion exchange resin (IER) collectors that are used to measure bulk (wet) deposition in interplant open spaces and throughfall (wet and dry) deposition under the dominant shrub.
\begin{itemize}
  \item run\_id - unique identifer of Lachat run provided primarily to allow users to associate blanks with samples
  \item field\_id - sample identifer
  \item collection\_date - date collectors are extracted from their deployment in the field
  \item notes - technician notes regarding conditions in the field or sample processing
  \item analyte\_name - nitrogen species measured (ammonium-nitrogen or nitrate-nitrogen)
  \item concentation - concentration of nitrogen species measured as ammonium-nitrogen or nitrate-nitrogen
  \item site\_code - site name abbreviation
\end{itemize}

### Fertilizer_Application
Catalog of amounts and timing of nitrogen and phosphorus fertilizer applications to nitrogen (N), phosphorus (P), and nitrogen+phosphorus (N+P) treatment plots - applications are delivered to the respective plot at the site that receives either N, P, or an N+P addition, except control sites (n = 2) that do not receive a fertilizer amendment.
\begin{itemize}
  \item site\_code - site name abbreviation
  \item application\_date - date of fertilizer application
  \item nitrogen - amount of ammonium-nitrate nitrogen fertilizer applied to 20mx20m nitrogen (N) and nitrogen+phosphorus (N+P) treatment plots
  \item phosphorus - amount of triple-super phosphate phosphorus fertilizer applied to 20mx20m phosphorus (P) and nitrogen+phosphorus (N+P) treatment plots
\end{itemize}

### Plant_Root_Simulator
Soil ion concentrations as determined with Plant Root Simulator (PRS®) probes (ion exchange resin membranes). Probes for the analyses of soil anions have a positively-charged membrane to simultaneously attract and adsorb all negatively-charged anions, such as nitrate (NO3-), phosphate (H2PO4-, HPO42-), and sulphate (SO42-), whereas cation probes have a negatively-charged membrane to simultaneously attract and adsorb all positively-charged cations, such as ammonium (NH4+), potassium (K+), calcium (Ca2+), and magnesium (Mg2+).
\begin{itemize}
  \item site\_code - site name abbreviation
  \item plot\_id - plot id number
  \item treatment\_code - treatment name abbreviation
  \item start\_date - date of probe deployment
  \item end\_date - date of probe extraction
  \item analyte - analyte measured
  \item final\_value - analyte concentration
  \item flag - data quality flag
  \item location\_within\_plot - habitat within plot where probes were deployed
  \item num\_cation\_probes - number of probes designated for analysis of cations recovered (of four)
  \item num\_anion\_probes - number of probes designated for analysis of anions recovered (of four)
  \item notes - notes pertaining to the deployment or analysis of the probes
\end{itemize}

### Stem_Growth
Biannual measures of stem growth on five Larrea tridentata study plants in Desert Fertilization experiment treatment and control plots.
\begin{itemize}
  \item site\_code - site name abbreviation
  \item plot\_id - plot id number
  \item treatment\_code - treatment name abbreviation
  \item scientific\_name - scientific name of plant measured
  \item shrub\_code - study plant identifier
  \item direction - direction or quadrant of plant in which measured stem is positioned
  \item pre\_date - date of initial or pre stem-length measurement
  \item post\_date - date of final or post stem-length
  \item post\_note - field note regarding final or post stem-length measurement
  \item post\_measurement - boolean indicator denoting whether the measurement is a pre or initial stem-length measurement (FALSE), or a final or post stem-length measurement (TRUE)
  \item stem\_length - length of stem
  \item stem\_comment - field note(s) regarding particular shrub or measurement
  \item plot\_comment - field note(s) regarding particular the plot or site
\end{itemize}

### Tissue_Chn
CHN (Carbon, Hydrogen, and Nitrogen) elemental analysis of Larrea tridentata leaf tissue and Pectocarya recurvata (whole plant) tissue collected from control plots at Desert Fertilization study sites.
\begin{itemize}
  \item site\_code - site name abbreviation
  \item plot\_id - plot id number
  \item treatment\_code - treatment name abbreviation
  \item sample\_date - date of leaf (Larrea) or plant (Pectocarya) collection
  \item season\_year - season and year of collection (for easier sorting)
  \item tissue\_type - source of tissue material
  \item analyte - element analyzed
  \item weight - weight of material used for analysis
  \item percent\_compostition - percent composition of analyte
  \item comment - comment about analytical measurement
\end{itemize}

### Tissue_Icp
Elemental composition of Larrea tridentata leaf tissue and Pectocarya recurvata (whole plant) tissue collected from control plots at Desert Fertilization study sites. Most analyses are by ICP-MS except Sulfur (S), which is typically analyzed by ICP-OES with the instrument type noted in the instrument field.
\begin{itemize}
  \item site\_code - site name abbreviation
  \item plot\_id - plot id number
  \item treatment\_code - treatment name abbreviation
  \item sample\_date - date of leaf (Larrea) or plant (Pectocarya) collection
  \item season\_year - season and year of collection (for easier sorting)
  \item tissue\_type - source of tissue material
  \item instrument - instrument employed for tissue analysis
  \item isotope\_element - isotope atomic mass and element symbol, except Sulfur (S; element symbol only)
  \item concentration - mass of isotope\_element per mass of leaf material
  \item source\_file - file name corresponding to raw data for the observation (raw data available as another entity in this data set)
\end{itemize}

## Code Definitions
### Site
\begin{itemize}
  \item DBG - core region: Desert Botanical Garden
  \item EME - west region: Estrella Mountain Regional Park east
  \item EMW - west region: Estrella Mountain Regional Park west
  \item LDP - east region: Lost Dutchman State Park
  \item MCN - east region: McDowell Mountain Regional north
  \item MCS - east region: McDowell Mountain Regional south
  \item MVP - core region: North Mountain
  \item PWP - core region: Piestewa Peak
  \item SME - core region: South Mountain Park east
  \item SMW - core region: South Mountain Park west
  \item SNE - west region: Sonoran Desert National Monument east
  \item SNW - west region: Sonoran Desert National Monument west
  \item SRR - east region: Salt River Recreation Area (Tonto NF)
  \item UMP - east region: Usery Mountain Regional Park
  \item WTM - west region: White Tanks Mountain Regional Park
\end{itemize} 

### Treatment
\begin{itemize}
  \item C1 - control plot 1
  \item N	- nitrogen amendment
  \item NP - nitrogen + phosphorus amendment
  \item P	- phosphorus amendment
\end{itemize}

### Location Within Plot (Biomass / Composition)
\begin{itemize}
  \item IP - subplot located in an interplant space	
  \item P	- subplot features a Larrea tridentata plant
\end{itemize}

### Cover Category
\begin{itemize}
  \item annual - observation referring to a type of plant
  \item plot\_characteristic - observation referring to characteristic of the plot other than cover of a specific type of annual plant
  \item record\_keeping - entry for record keeping, typically whether the plot was surveyed
\end{itemize}

### Location Within Plot (Simulator)
\begin{itemize}
  \item between\_plant - probes buried between plant canopies
  \item blank - field blank: probe stored in a clean container under refrigeration during the deployment period
  \item blk\_washed - field blank (washed): probe stored in a clean container under refrigeration during the deployment period, and washed as per sample probes
  \item under\_plant - probes buried beneath the canopy of Larrea tridentata
\end{itemize}

### Analyte
\begin{itemize}
  \item Al - aluminum	
  \item B	- boron	
  \item Ca - calcium	
  \item Cd - cadmium	
  \item Cu - copper	
  \item Fe - iron	
  \item K	- potassium	
  \item Mg - magnesium	
  \item Mn - manganese	
  \item NH4-N	- ammonium-nitrogen	
  \item NO3-N	- nitrate-nitrogen	
  \item P	- phosphorus	
  \item Pb - lead	
  \item S	- sulfur	
  \item Total-N	- sum of NO3-N and NH4-N	
  \item Zn - zinc
\end{itemize}

### Shrub
\begin{itemize}
  \item L1 - Larrea tridentata study plant \#1	
  \item L2 - Larrea tridentata study plant \#2	
  \item L3 - Larrea tridentata study plant \#3	
  \item L4 - Larrea tridentata study plant \#4	
  \item L5 - Larrea tridentata study plant \#5
\end{itemize}

### Tissue Type
\begin{itemize}
  \item Larrea tridentata - Larrea tridentata leaf tissue	
  \item Pectocarya recurvata - Pectocarya recurvata whole-plant tissue	
\end{itemize}

### Analyte (Tissue_chn)
\begin{itemize}
  \item Carbon \% - percent carbon	
  \item Hydrogen \% - percent hydrogen	
  \item Nitrogen \% - percent nitrogen
\end{itemize}

### Instrument
\begin{itemize}
  \item ICP-MS - Inductively Coupled Plasma Mass Spectrometer	
  \item ICP-OES - Inductively Coupled Plasma Optical Emission Spectrometer
\end{itemize}